---
title: "Animating Stacked Maps"
vignette: >
  %\VignetteIndexEntry{Animating Stacked Maps}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
---

# Introduction

One of the most visually striking uses of stacked maps is to animate the "infolding" or "revealing" of the layers. `{layer}` provides a high-level API to create these animations with a single function call, while remaining fully compatible with `{gganimate}` for more advanced scenarios.

# Quick Start: High-Level Animation

The easiest way to animate a stack is using `animate_tilt_stack()`. It takes a `tilt_stack` object and generates a `gganimate` object.

There are two primary animation types:

1.  **`"unfold"` (Default):** All layers transition from a flat 2D state to their final 3D positions simultaneously.
2.  **`"reveal"`:** Layers appear and transition into place one by one.

```{r}
library(layer)
library(sf)
library(ggplot2)
library(gganimate)

# Load base data
data("landscape_1")
data("landscape_2")
data("landscape_3")

# Define a standard stack
stack <- tilt_stack(y_shift_step = 50) |>
  tilt_layer(landscape_1, palette = "bilbao", label = "Bottom") |>
  tilt_layer(landscape_2, palette = "mako", label = "Middle") |>
  tilt_layer(landscape_3, palette = "rocket", label = "Top")

# Create an "accordion" unfold animation
# mirai::daemons(0)
# mirai::daemons(4)
anim <- animate_tilt_stack(stack, type = "unfold", n_frames = 10)

# Render the animation
# gganimate::anim_save("unfold_animation.gif", anim, fps = 25)
```

# Advanced: Custom Movement & Matching Data

While `animate_tilt_stack()` handles the most common use cases, you can also build custom animations by manually tilting your data. This is particularly useful when you have moving objects (like vehicles or wildlife) that need to stay "pinned" to a specific layer.

## Setup for Custom Animation

We'll generate some fake "routes" and "buses" to animate.

```{r}
library(layer)
library(sf)
library(ggplot2)
library(gganimate)

# Load base data
data("landscape_1")
data("landscape_2")

# Create a fake public transport route
route_line <- st_linestring(matrix(c(10, 10, 90, 90), ncol = 2, byrow = TRUE))
route_sf <- st_sf(route_id = "Route A", geometry = st_sfc(route_line))

# Generate 20 frames of movement along the route
n_frames <- 20
bus_pts <- st_line_sample(route_line, n = n_frames, type = "regular") |>
  st_cast("POINT")

# Combine into a data frame with timestamps
buses <- st_sf(
  route_id = "Route A",
  time = seq(
    as.POSIXct("2021-11-11 07:00:00"),
    as.POSIXct("2021-11-11 08:00:00"),
    length.out = n_frames
  ),
  geometry = bus_pts
)
```

# Building the Static Foundation

We use the declarative `tilt_stack()` API to build the 3D environment. We'll place the transport routes on the top layer (`y_shift = 50`).

```{r}
# Define the stack
# We use y_shift_step = 25, so Layer 1 is at 0, Layer 2 is at 25, Layer 3 is at 50
map_stack <- tilt_stack(y_shift_step = 25) |>
  tilt_layer(landscape_1, palette = "bilbao", label = "Terrain") |>
  tilt_layer(landscape_2, palette = "mako", label = "Population") |>
  tilt_layer(route_sf, color = "white", size = 1.5, label = "Transport Network")

# Render the static base
base_plot <- plot_tilt_stack(map_stack)
```

# Preparing the Animated Points

To animate the buses, we need to tilt their coordinates so they perfectly align with the top layer of our stack. We can now use the `tilt_match()` utility which automatically handles the spatial transformations based on our stack definition.

```{r}
# Tilt the buses to match the 3rd layer of our stack
buses_tilted <- tilt_match(buses, map_stack, layer = 3)

# Extract coordinates into a regular data frame for gganimate performance
buses_df <- data.frame(
  route_id = buses$route_id,
  time = buses$time,
  x = st_coordinates(buses_tilted)[, 1],
  y = st_coordinates(buses_tilted)[, 2]
)
```

# Assembling the Animation

Finally, we append the moving points and `gganimate` instructions to our static base plot.

```{r}
animated_plot <- base_plot +
  # Add a new color scale for the buses
  ggnewscale::new_scale_color() +
  geom_point(data = buses_df, aes(x = x, y = y), color = "red", size = 4) +

  # gganimate specifications
  labs(title = 'Time: {frame_time}') +
  transition_time(time) +
  shadow_wake(wake_length = 0.1) +
  ease_aes('linear')

# Render and save
# anim_save("stacked_animation.gif", animated_plot)
```

# Why this works

Because `plot_tilt_stack()` returns a standard `ggplot` object, we can continue to add layers and scales to it. By manually tilting the "dynamic" data using the same shift parameters as our "static" layers, we ensure everything remains perfectly aligned in the 3D perspective.
