---
title: "Get Started with {layer}"
vignette: >
  %\VignetteIndexEntry{Get Started with {layer}}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
---

# Introduction

`{layer}` is an R package designed to simplify the creation of stacked tilted maps, commonly used in scientific publications to visualize spatial correlations across different environmental layers.

The package provides a declarative API to easily build map stacks, handle spatial transformations (tilting, shifting, and stretching), and automatically generate connecting lines between layers with perfect 3D occlusion.

# Installation

You can install the development version of `{layer}` from GitHub with:

```{r}
# install.packages("remotes")
remotes::install_github("marcosci/layer")
```

# Basic Usage

The core of the package is the `tilt_stack()` system. You initialize a stack, add map layers with `tilt_layer()`, and optionally add connecting pins with `tilt_connector()`.

Load relevant packages:

```{r}
library(layer)
library(sf)
library(ggplot2)
```

## Simple Stack Example

Here is how to create a basic 3-layer stack using the built-in landscape data:


```{r}
# Initialize a stack with a global vertical step
a
```

## How it Works

- **`tilt_stack()`**: Defines global defaults for the stack, such as the shift step between layers and default labels positioning.
- **`tilt_layer()`**: Adds a raster or vector layer. It inherits global spatial parameters unless overridden. The `label` argument automatically places text next to the layer in 3D space.
- **`tilt_connector()`**: Takes a set of points and automatically calculates their positions on every layer in the stack, drawing vertical connecting lines that respect the 3D occlusion. By default, lines are interleaved between layers (`on_top = FALSE`), but can be forced to the front with `on_top = TRUE`.

### Global vs. Local Connectors

The `tilt_connector()` function is "position-aware" within your pipeline:

1.  **Global Connectors:** If added at the very beginning or end of your stack, the connector pierces through **every** layer gap.
2.  **Local Connectors:** If added between two specific `tilt_layer()` calls, it only draws connecting lines for that specific gap.

```{r}
# Generate some sample points
points_global <- st_sf(id = 1, geometry = st_sfc(st_point(c(25, 25))))
points_local <- st_sf(id = 2, geometry = st_sfc(st_point(c(75, 75))))

# Example of mixed connectors
tilt_stack(x_shift_step = 50, y_shift_step = 50) |>
  # Global: Pierces the entire stack
  tilt_connector(points_global, color = "black", size = 1.5) |>

  tilt_layer(landscape_1, label = "Bottom") |>

  # Local: Only connects the Bottom to the Middle layer
  tilt_connector(points_local, color = "red", size = 1.5) |>

  tilt_layer(landscape_2, label = "Middle") |>
  tilt_layer(landscape_3, label = "Top")
``
`

# Advanced: Reproducing Complex Figures

The package is powerful enough to recreate complex, multi-sourced spatial visualizations. Because of the built-in S3 print method, simply running the pipeline in your console will automatically render the final 3D visualization. 

Note that if you wish to save the resulting `
ggplot
` object to a variable without printing it immediately, you can still use the explicit `
plot_tilt_stack()
` function:

`
``
{
  r
}
# Initialize stack with vertical separation
my_plot <- tilt_stack(
  x_shift_step = 0,
  y_shift_step = 25,
  label_color = "gray40"
) |>
  # Terrain raster
  tilt_layer(landscape_1, palette = "YlOrRd", label = "Terrain") |>
  # ... (other layers)
  plot_tilt_stack() # Explicit plotting
```

```{r}
# Fully declarative pipeline (auto-renders)
tilt_stack(x_shift_step = 0, y_shift_step = 25, label_color = "gray40") |>
  tilt_layer(landscape_1, palette = "YlOrRd", label = "Terrain") |>
  tilt_layer(landscape_2, palette = "cividis", label = "Population") |>
  tilt_layer(landscape_1, fill = NA, color = "gray50", label = "Boundaries") |>
  tilt_layer(landscape_points, color = "black", y_shift_rel = 50) |>
  tilt_layer(landscape_3, palette = "viridis", label = "Accessibility")
```

# Legacy API (Deprecated)

Before the introduction of the `tilt_stack` system, users had to manually tilt each layer using `tilt_map()` and then combine them in a list for `plot_tiltedmaps()`. This approach is now **deprecated** as it requires manual coordinate tracking and doesn't support automated 3D occlusion for connectors.

```r
# The old, manual way (deprecated)
l1 <- tilt_map(landscape_1)
l2 <- tilt_map(landscape_2, x_shift = 25, y_shift = 50)
plot_tiltedmaps(list(l1, l2), palette = c("bilbao", "mako"))
```

# Conclusion

The `{layer}` package removes the manual math usually required for 3D map stacks in R. By using a declarative pipeline, your code stays readable, and the system automatically handles shears, rotations, and perfect 3D occlusion for you.
